Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-19 numerical-amr.tex
Formatting police passing through.

One of the big advantages a flux-conservative methods is, of course, their ability to maintain conserved-quantities to machine precision.  Unfortunately, the projection step upsets this flux conservation since, at the boundary of a refined region, the coarse and fine cells on either side of the border are updated with different estimates of the flux across the boundary.  In one case this comes from the coarse grid computation, and in the other from the fine grid solution.  Conservation can be restored by correcting all coarse cells which lie at the boundary of a fine region
with the difference between the coarse and fine estimates of the flux across the boundary:
\begin{equation}
q^{\rm coarse} = \tilde{q}^{\rm coarse} - \Delta t \left( F^{\rm coarse} - \sum_{j,k} F^{fine}_{j,k} \right)
\end{equation}
where $F$ is the flux of quantity $q_i$ in the $i$-direction through the cell interface.   We use $\tilde{q}$ to indicate the uncorrected quantity on the coarse grid and the sum is over the $r^{d-1}$ fine cell interfaces in the perpendicular dimensions which share a face with the coarse cell being corrected.  This corrected is carried out for all coarse cells which share a face with a refined cell (and are not themselves covered with more refined cells).

@@

\item{\em Jeans Length.}  Following \cite{Truelove98}, the code can be forced to refine the Jeans length by a fixed number of cells, or, in more specifically, whenever the following criterion is not met:
\begin{equation}
\Delta x < \left( \frac{\pi k_B T}{J^2 G \rho m_H} \right)^{1/2}
\end{equation}
where $J$ is the required number of cells per Jeans length (4 by default).

\item{\em Cooling time.}  Often in cosmological simulations, cooling is a rapid process and we do not necessarily want to be limited by the cooling time.  However, in some cases it is desirable to resolve the cooling process, in which case we flag on cells which have a cooling time shorter than their sound crossing time.  The cooling time is given by $t_{\rm cool} = 1.5 kT / (n \Lambda(T)$ and the sound crossing time is $t_{\rm cross} = \Delta x / c_s$, where $c_s$ is the sound speed.

@@

\item{\em Optical depth.} This refinement criterion ensures that cells on the highest level of refinement have an HI optical depth lower than unity.  The optical depth of a cell is estimated to be:

\begin{equation}
\tau = \sigma_{HI} n_{HI} \Delta x 
\end{equation}

where $\sigma_{HI}$ is the neutral atomic hydrogen absorption cross section, $n_{HI}$ is the proper HI number density, and $\Delta x$ is the cell width.  If $\tau > 1$, the cell is refined further, up to the maximum level of refinement.

>>>>

One of the big advantages a flux-conservative methods is, of course, their ability to maintain conserved-quantities to machine precision.  Unfortunately, the projection step upsets this flux conservation since, at the boundary of a refined region, the coarse and fine cells on either side of the border are updated with different estimates of the flux across the boundary.  In one case this comes from the coarse grid computation, and in the other from the fine grid solution.  Conservation can be restored by correcting all coarse cells which lie at the boundary of a fine region
with the difference between the coarse and fine estimates of the flux across the boundary:
\begin{equation}
  q^{\rm coarse} = \tilde{q}^{\rm coarse} - \Delta t \left( F^{\rm
      coarse} - \sum_{j,k} F^{\rm fine}_{j,k} \right)
\end{equation}
where $F$ is the flux of quantity $q_i$ in the $i$-direction through the cell interface.   We use $\tilde{q}$ to indicate the uncorrected quantity on the coarse grid and the sum is over the $r^{d-1}$ fine cell interfaces in the perpendicular dimensions which share a face with the coarse cell being corrected.  This corrected is carried out for all coarse cells which share a face with a refined cell (and are not themselves covered with more refined cells).

@@

\item{\em Jeans Length.}  Following \cite{Truelove98}, the code can be forced to refine the Jeans length by a fixed number of cells, or, in more specifically, whenever the following criterion is not met:
\begin{equation}
\Delta x < \left( \frac{\pi k_B T}{J^2 G \rho \mh} \right)^{1/2}
\end{equation}
where $J$ is the required number of cells per Jeans length (4 by default).

\item{\em Cooling time.}  Often in cosmological simulations, cooling is a rapid process and we do not necessarily want to be limited by the cooling time.  However, in some cases it is desirable to resolve the cooling process, in which case we flag on cells which have a cooling time shorter than their sound crossing time.  The cooling time is given by $t_{\rm cool} = 1.5 kT / n \Lambda(T)$ and the sound crossing time is $t_{\rm cross} = \Delta x / c_s$, where $c_s$ is the sound speed.

@@

\item{\em Optical depth.} This refinement criterion ensures that cells on the highest level of refinement have an \ion{H}{1} optical depth lower than unity.  The optical depth of a cell is estimated to be:

\begin{equation}
\tau = \sigma_{\rm HI} n_{\rm HI} \Delta x 
\end{equation}

where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width.  If $\tau > 1$, the cell is refined
further, up to the maximum level of refinement.

### 2013-03-19 numerical-analysis.tex
Formatting police passing through.

\begin{eqnarray}
\label{eqn.drifttrace}
x^{n+1/2} & = & x^n + (\Delta t/2) v^{interp,n} \nonumber \\
v^{n+1} & = & v^{interp,n+1} \\
x^{n+1} & = & x^{n+1/2} + (\Delta t/2) v^{n+1} \nonumber
\end{eqnarray}

This is essentially a drift-kick-drift particle update from time $n$
to time $n+1$ -- however, instead of computing an acceleration at the
half-timestep $t^n + \Delta t/2$ (as is done for massive particles --
see Equation~\ref{eqn.driftkick}), the particle's velocity is updated
both at the beginning of each timestep and at the half-timestep
by linearly interpolating the cell-centered baryon velocity to the
position of the particle ($v^{interp}$), and assigning it to that value.

>>>>

\begin{eqnarray}
\label{eqn.drifttrace}
x^{n+1/2} & = & x^n + (\Delta t/2) v^{\rm interp,n} \nonumber \\
v^{n+1} & = & v^{\rm interp,n+1} \\
x^{n+1} & = & x^{n+1/2} + (\Delta t/2) v^{n+1} \nonumber
\end{eqnarray}

This is essentially a drift-kick-drift particle update from time $n$
to time $n+1$ -- however, instead of computing an acceleration at the
half-timestep $t^n + \Delta t/2$ (as is done for massive particles --
see Equation~\ref{eqn.driftkick}), the particle's velocity is updated
both at the beginning of each timestep and at the half-timestep
by linearly interpolating the cell-centered baryon velocity to the
position of the particle ($v^{\rm interp}$), and assigning it to that value.

### 2013-03-19 numerical-overview.tex
Formatting police passing through.

Several processes are neglected.  The deuterium atom and its processes are 
completely ignored, which may have some effect.  Recent work shows that HD is
a more effective coolant than previously thought~\citep{2005MNRAS.361..850L}.  However,
the fractional abundance of HD is so low that under circumstances relevant to
the formation of a Population III star in an un-ionized region it should be
sub-dominant.  However, the enhanced electron fraction in fossil 
HII regions (as discussed later in this thesis) could result in the HD molecule
becoming a dominant cooling mechanism at relatively low ($\sim$ few hundred K)
temperatures, and could potentially cool the gas down to below $100$~K, which can
enhance fragmentation and could have important consequences for the IMF of 
primordial stars forming in a relic HII region~\citep{2005MNRAS.364.1378N}.

>>>>

Several processes are neglected.  The deuterium atom and its processes are 
completely ignored, which may have some effect.  Recent work shows that HD is
a more effective coolant than previously thought~\citep{2005MNRAS.361..850L}.  However,
the fractional abundance of HD is so low that under circumstances relevant to
the formation of a Population III star in an un-ionized region it should be
sub-dominant.  However, the enhanced electron fraction in fossil 
\ion{H}{2} regions (as discussed later in this thesis) could result in the HD molecule
becoming a dominant cooling mechanism at relatively low ($\sim$ few hundred K)
temperatures, and could potentially cool the gas down to below $100$~K, which can
enhance fragmentation and could have important consequences for the IMF of 
primordial stars forming in a relic \ion{H}{2} region~\citep{2005MNRAS.364.1378N}.

### 2013-03-19 numerical-radiation-fld.tex
Formatting police passing through.

Enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities, 
\begin{eqnarray}
  \label{eq:fld_radiation}
  \partial_t E_r + \frac1a \nabla\cdot\left(E_r \vec{v}_b\right) &=& 
    \nabla\cdot\left(D\nabla E_r\right) -
    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
  \label{eq:fld_heating}
  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
  \label{eq:fld_chemistry}
  \partial_t {\mathrm n}_i + 
    \frac1a \nabla\cdot\left({\mathrm n}_i \vec{v}_b\right) &=& 
    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
\end{eqnarray}
where $E_r$ is a grey radiation energy density, $e_c$ is the
internal energy correction due to photo-heating and chemical cooling,
and ${\mathrm n}_i$ correspond to the chemical number densities of
HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
number density.  Here, we define the grey radiation energy density
through first assuming a fixed frequency spectrum, i.e.
$E_{\nu}(\nu,\vec{x},t) = \tilde{E}_r(\vec{x},t) \chi(\nu)$, and then
defining the integrated quantity
\begin{equation}
\label{eq:grey_radiation_energy}
   E_r(\vec{x},t) \equiv \int_{\nu_0}^{\infty}
   E_{\nu}(\nu,\vec{x},t)\,\mathrm d\nu \  = \ 
   \tilde{E}_r(\vec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
\end{equation}
In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
opacity, $\eta$ is the field of radiation sources, $G$ provides the
radiation induced photo-heating, $\Lambda$ provides the chemical
cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

>>>>

Enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities, 
\begin{eqnarray}
  \label{eq:fld_radiation}
  \partial_t E_r + \frac1a \nabla\cdot\left(E_r \vec{v}_b\right) &=& 
    \nabla\cdot\left(D\nabla E_r\right) -
    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
  \label{eq:fld_heating}
  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
  \label{eq:fld_chemistry}
  \partial_t {\mathrm n}_i + 
    \frac1a \nabla\cdot\left({\mathrm n}_i \vec{v}_b\right) &=& 
    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
\end{eqnarray}
where $E_r$ is a grey radiation energy density, $e_c$ is the internal
energy correction due to photo-heating and chemical cooling, and
${\mathrm n}_i$ correspond to the chemical number densities of
\ion{H}{1}, \ion{H}{2}, \ion{He}{1}, \ion{He}{2} and \ion{He}{3}, and
${\mathrm n}_e$ is the electron number density.  Here, we define the
grey radiation energy density through first assuming a fixed frequency
spectrum, i.e.  $E_{\nu}(\nu,\vec{x},t) = \tilde{E}_r(\vec{x},t)
\chi(\nu)$, and then defining the integrated quantity
\begin{equation}
\label{eq:grey_radiation_energy}
   E_r(\vec{x},t) \equiv \int_{\nu_0}^{\infty}
   E_{\nu}(\nu,\vec{x},t)\,\mathrm d\nu \  = \ 
   \tilde{E}_r(\vec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
\end{equation}
In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
opacity, $\eta$ is the field of radiation sources, $G$ provides the
radiation induced photo-heating, $\Lambda$ provides the chemical
cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

### 2013-03-19 numerical-timestep.tex
Formatting police passing through.

\begin{equation}
\Delta t_{hydro} = min \left( \kappa_{hydro} \frac{a \Delta x}{c_{s} + |v_x|} \right)_L ,
\label{eqn:dthydro}
\end{equation}

\begin{equation}
\Delta t_{MHD} = min \left( \kappa_{MHD} \frac{a \Delta x}{v_{f} + |v_x|} \right)_L ,
\label{eqn:dtMHD}
\end{equation}

\begin{equation}
\Delta t_{dm} = min \left(\kappa_{dm} \frac{a \Delta x}{v_{dm,x}} \right)_L ,
\label{eqn:dtdarkmatter}
\end{equation}

\begin{equation}
\Delta t_{accel} = min \left( \sqrt{\frac{\Delta x}{\vec{g}}} \right)_L ,
\label{eqn:dtaccel}
\end{equation}

\begin{equation}
\Delta t_{rad} = min \left(  \sqrt{\frac{\Delta x}{\vec{a}_{rad}} \right)_L,
\label{eqn:dtrad}
\end{equation}

\begin{equation}
\Delta t_{cond} = min \left(  \frac{ \kappa_{cond}}{f_{Sp}} \frac{\Delta x^2
    n_b}{\kappa(T)} \right)_L,
\label{eqn:dtcond}
\end{equation}

\begin{equation}
\Delta t_{exp} = f_{exp} \left( \frac{a}{\dot{a}} \right) ,
\label{eqn:dtexpand}
\end{equation}

@@

 In equations~\ref{eqn:dthydro}-\ref{eqn:dtcond}, the $min ( \ldots
)_L$ formalism means that this value is calculated for all cells on a
given level L and the minimum overall value is taken as the timestep.
Equation~\ref{eqn:dthydro} ensures that all cells satisfy the
Courant-Freidrichs-Levy (CFL) condition for accuracy and stability of
an explicit finite difference discretization of the Euler equations.
Effectively this condition forces the timestep to be small enough such
that any changes in the fluid propagate less than a single grid
spacing, $\Delta x$.  In this equation, $\kappa_{hydro}$ is a
numerical constant with a value of $0 < \kappa_{hydro} \leq 1$ (with a
typical value of $\kappa_{hydro} \sim 0.3-0.5$) that ensures that the
CFL condition is always met, and $c_s$ and $v_x$ are the sound speed
and peculiar baryon velocity in a given cell.

Equation ~\ref{eqn:dthydro} is valid for one dimension.  For 2 or 3
dimensions, it was shown by \cite{Godunov1959}  that using the
harmonic average of the timestep found along each of the coordinate
axes yields a maximum $\kappa_{hydro} = 0.8$.  So letting $\Delta
t_x$, $\Delta t_y$, and $\Delta t_z$ be the analogues of
eqn. \ref{eqn:dthydro} along the $x,y$ and $z$ axes, 
\begin{equation}
\Delta t_{hydro} = min ( \kappa_{hydro} /( \frac{1}{\Delta t_x}
+\frac{1}{\Delta t_y} + \frac{1}{\Delta t_z} ) )_L
\end{equation}

@@

Equation~\ref{eqn:dtMHD} is only enforced when the equations of
magnetohydrodynamics are being solved, and is directly analogous to
Equation~\ref{eqn:dthydro} in that it ensures that the CLF condition
is being enforced at all times.  In this equation, $\kappa_{MHD}$ is a
numerical constant with a value of $0 < \kappa_{MHD} \leq 1$ (with a
typical value of $\kappa_{MHD} \sim 0.5$) that ensures that the CFL
condition is always met, and $v_f$ and $v_x$ are the ``fast wave
speed'' and peculiar baryon velocity in a given cell.  The fast wave
speed comes from a stability analysis of the MHD equations, and is
expressed as:

\begin{equation}
v_f = \sqrt{ \frac{1}{2} \left(  v_A^2 + c_s^2 + \sqrt{(v_A^2 +
      c_s^2)^2 - 4 v_A^2 c_s^2}  \right)  },
\label{eqn:vfastmhd}
\end{equation}

Where $c_s$ is the sound speed and $v_A$ is the Alfven speed, calculated
as $v_A = \sqrt{B^2/\rho_B}$ in units where $\mu_0 = 1$.

Equation~\ref{eqn:dtdarkmatter} is analogous to
Equation~\ref{eqn:dthydro} and ensures accuracy in the N-body solver
by requiring that no dark matter particle travels more than one cell
width.  The parameter $\kappa_{dm}$ is analogous to
$\kappa_{dthydro}$, with an identical range of values.
Equation~\ref{eqn:dtexpand} limits the timestep such that the
expansion parameter a can only change by a fractional amount of
$f_{exp} = \Delta a/a$, where $f_{exp}$ is a user-defined parameter
and has typical values of $f_{exp} = 0.01-0.02$.  This is required for
the stability of the PPM algorithm in comoving coordinates, and
typically limits the timestep only at high redshifts when densities
are relatively homogeneous.

Equation~\ref{eqn:dtaccel} and~\ref{eqn:dtrad} are supplementary to equation~\ref{eqn:dthydro} in that they
take into account the possibility of large accelerations due to either
gravity (Eqn.~\ref{eqn:dtaccel} or radiation pressure
(Eqn.~\ref{eqn:dtrad}), which may cause numerical 
instabilities by violating the Courant condition.  In Equation~\ref{eqn:dtaccel}, $\vec{g}$ is the
gravitational acceleration in each cell on level L.  In
Equation~\ref{eqn:dtrad}, $\vec{a}_{rad}$ is the estimated
acceleration due to radiation pressure in each cell on level L,
defined as

@@

Equation~\ref{eqn:dtcond} is the CFL condition for an explicit
solution to the equation of heat conduction.  In this expression,
$n_b$ is the baryon number density, defined as $n_b = \rho_b / \mu
m_p$, where $\mu$ is the mean molecular weight, $\kappa(T)$ is the
Spitzer thermal conductivity, and f$_{Sp}$ is a user-defined
conduction suppression factor whose value must be f$_{Sp} \leq 1$.  Both are defined in
Section~\ref{sec.num.conductions}.  \kappa_{cond} is a prefactor whose
value must be $0 < \kappa_{cond}  < 0.5$, and is exactly 0.5 for the
implemention in Enzo.
From a practical
perspective, it is useful to note that, unlike other timestep criteria
discussed above (which effectively scale as $\Delta x / \sqrt{T}$,
the timestep criterion due to thermal conduction scales as $\Delta x^2
/ T^{2.5}$, which can result in a rapid decrease in timestep in
regions of high resolution and/or temperature.

Finally, Equation~\ref{eqn:dtexpand} is a cosmological constraint that
limits the timestep so that the simulated universe only expands by
some fractional amount, f$_{exp}$, during a single step.  In this
equation, $a$ and $\dot{a}$ refer to the scale factor of the universe
and its rate of change, respectively.  This
criteria is necessary because the expansion of the universe and its
first derivative with respect to time both appear in the equations of cosmological (magneto)hydrodynamics and particle motion,
and some limit is required in order to model the change of energy and
momentum due to expansion with a reasonable level of accuracy.  In
reality, this is only a significant limitation at very high redshift;
at later times, other timestep criteria are much more important.

>>>>

\begin{equation}
\Delta t_{\rm hydro} = \min \left( \kappa_{\rm hydro} \frac{a \Delta x}{c_{s} + |v_x|} \right)_L ,
\label{eqn:dthydro}
\end{equation}

\begin{equation}
\Delta t_{\rm MHD} = \min \left( \kappa_{\rm MHD} \frac{a \Delta x}{v_{f} + |v_x|} \right)_L ,
\label{eqn:dtMHD}
\end{equation}

\begin{equation}
\Delta t_{\rm dm} = \min \left(\kappa_{\rm dm} \frac{a \Delta x}{v_{dm,x}} \right)_L ,
\label{eqn:dtdarkmatter}
\end{equation}

\begin{equation}
\Delta t_{\rm accel} = \min \left( \sqrt{\frac{\Delta x}{\vec{g}}} \right)_L ,
\label{eqn:dtaccel}
\end{equation}

\begin{equation}
\Delta t_{\rm rad} = \min \left(  \sqrt{\frac{\Delta x}{\vec{a}_{rad}}} \right)_L,
\label{eqn:dtrad}
\end{equation}

\begin{equation}
\Delta t_{\rm cond} = \min \left(  \frac{ \kappa_{\rm cond}}{f_{\rm Sp}} \frac{\Delta x^2
    n_b}{\kappa(T)} \right)_L,
\label{eqn:dtcond}
\end{equation}

\begin{equation}
\Delta t_{\rm exp} = f_{\rm exp} \left( \frac{a}{\dot{a}} \right) ,
\label{eqn:dtexpand}
\end{equation}

@@

In equations~\ref{eqn:dthydro}-\ref{eqn:dtcond}, the $\min ( \ldots
)_L$ formalism means that this value is calculated for all cells on a
given level L and the minimum overall value is taken as the timestep.
Equation~\ref{eqn:dthydro} ensures that all cells satisfy the
Courant-Freidrichs-Levy (CFL) condition for accuracy and stability of
an explicit finite difference discretization of the Euler equations.
Effectively this condition forces the timestep to be small enough such
that any changes in the fluid propagate less than a single grid
spacing, $\Delta x$.  In this equation, $\kappa_{\rm hydro}$ is a
numerical constant with a value of $0 < \kappa_{\rm hydro} \leq 1$
(with a typical value of $\kappa_{\rm hydro} \sim 0.3-0.5$) that
ensures that the CFL condition is always met, and $c_s$ and $v_x$ are
the sound speed and peculiar baryon velocity in a given cell.

Equation ~\ref{eqn:dthydro} is valid for one dimension.  For 2 or 3
dimensions, it was shown by \cite{Godunov1959}  that using the
harmonic average of the timestep found along each of the coordinate
axes yields a maximum $\kappa_{\rm hydro} = 0.8$.  So letting $\Delta
t_x$, $\Delta t_y$, and $\Delta t_z$ be the analogues of
eqn. \ref{eqn:dthydro} along the $x,y$ and $z$ axes, 
\begin{equation}
  \Delta t_{\rm hydro} = \min \left( \frac{\kappa_{\rm hydro}} {1/\Delta t_x
  +1/\Delta t_y + 1/\Delta t_z} \right)_L
\end{equation}

@@

Equation~\ref{eqn:dtMHD} is only enforced when the equations of
magnetohydrodynamics are being solved, and is directly analogous to
Equation~\ref{eqn:dthydro} in that it ensures that the CFL condition
is being enforced at all times.  In this equation, $\kappa_{\rm MHD}$ is a
numerical constant with a value of $0 < \kappa_{\rm MHD} \leq 1$ (with a
typical value of $\kappa_{\rm MHD} \sim 0.5$) that ensures that the CFL
condition is always met, and $v_f$ and $v_x$ are the ``fast wave
speed'' and peculiar baryon velocity in a given cell.  The fast wave
speed comes from a stability analysis of the MHD equations, and is
expressed as:
%
\begin{equation}
v_f = \sqrt{ \frac{1}{2} \left(  v_A^2 + c_s^2 + \sqrt{(v_A^2 +
      c_s^2)^2 - 4 v_A^2 c_s^2}  \right)  },
\label{eqn:vfastmhd}
\end{equation}
%
where $c_s$ is the sound speed and $v_A$ is the Alfven speed, calculated
as $v_A = \sqrt{B^2/\rho_B}$ in units where $\mu_0 = 1$.

Equation~\ref{eqn:dtdarkmatter} is analogous to
Equation~\ref{eqn:dthydro} and ensures accuracy in the N-body solver
by requiring that no dark matter particle travels more than one cell
width.  The parameter $\kappa_{\rm dm}$ is analogous to $\kappa_{\rm
  hydro}$, with an identical range of values.
Equation~\ref{eqn:dtexpand} limits the timestep such that the
expansion parameter a can only change by a fractional amount of
$f_{\rm exp} = \Delta a/a$, where $f_{\rm exp}$ is a user-defined
parameter and has typical values of $f_{\rm exp} = 0.01-0.02$.  This
is required for the stability of the PPM algorithm in comoving
coordinates, and typically limits the timestep only at high redshifts
when densities are relatively homogeneous.

Equation~\ref{eqn:dtaccel} and~\ref{eqn:dtrad} are supplementary to equation~\ref{eqn:dthydro} in that they
take into account the possibility of large accelerations due to either
gravity (Eqn.~\ref{eqn:dtaccel} or radiation pressure
(Eqn.~\ref{eqn:dtrad}), which may cause numerical 
instabilities by violating the Courant condition.  In Equation~\ref{eqn:dtaccel}, $\vec{g}$ is the
gravitational acceleration in each cell on level L.  In
Equation~\ref{eqn:dtrad}, $\vec{a}_{\rm rad}$ is the estimated
acceleration due to radiation pressure in each cell on level L,
defined as

@@

Equation~\ref{eqn:dtcond} is the CFL condition for an explicit
solution to the equation of heat conduction.  In this expression,
$n_b$ is the baryon number density, defined as $n_b = \rho_b / \mu
m_p$, where $\mu$ is the mean molecular weight, $\kappa(T)$ is the
Spitzer thermal conductivity, and f$_{Sp}$ is a user-defined
conduction suppression factor whose value must be f$_{Sp} \leq 1$.  Both are defined in
Section~\ref{sec.num.conductions}.  $\kappa_{\rm cond}$ is a prefactor whose
value must be $0 < \kappa_{\rm cond}  < 0.5$, and is exactly 0.5 for the
implemention in Enzo.
From a practical
perspective, it is useful to note that, unlike other timestep criteria
discussed above (which effectively scale as $\Delta x / \sqrt{T}$,
the timestep criterion due to thermal conduction scales as $\Delta x^2
/ T^{2.5}$, which can result in a rapid decrease in timestep in
regions of high resolution and/or temperature.

Finally, Equation~\ref{eqn:dtexpand} is a cosmological constraint that
limits the timestep so that the simulated universe only expands by
some fractional amount, f$_{\rm exp}$, during a single step.  In this
equation, $a$ and $\dot{a}$ refer to the scale factor of the universe
and its rate of change, respectively.  This
criteria is necessary because the expansion of the universe and its
first derivative with respect to time both appear in the equations of cosmological (magneto)hydrodynamics and particle motion,
and some limit is required in order to model the change of energy and
momentum due to expansion with a reasonable level of accuracy.  In
reality, this is only a significant limitation at very high redshift;
at later times, other timestep criteria are much more important.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "ms"
%%% End:

### 2013-03-19 overview.tex
Formatting police passing through.

\begin{equation}
\mathbb{F} = -\kappa_{sp} \nabla T
\end{equation}

Where $\kappa_{sp}$ is the Spitzer conduction coefficient
\citep{1962pfig.book.....S}.  Saturation of the heat flux in
high-temperature, low-density regimes (such as the intracluster medium
in galaxy clusters) is taken into account.  Thermal conduction in a
plasma can be strongly affected by the presence of magnetic field
lines, which may strongly suppress heat flow perpendicular to the
magnetic field.  In that case, we allow for heat transport only along
the magnetic field lines in simulations that include this physics, as follows:

\begin{equation}
\mathbb{F} = -\kappa_{sp} \hat{b} (\hat{b} \cdot \nabla T)
\end{equation}

@@

In addition to this, \enzo\ includes the capability of following up to 12 chemical species using a non equilibrium-solver.  The species can be turned on in sets, with the simplest model including just H, H$^+$, He, He$^+$, He$^++$, and $e^-$, and more complete models adding first species important for gas-phase molecular hydrogen formation: $H^-$, H$_2$ and H$_2^+$, and then $HD$ formation: $HD$, $D$, $D+$.  The cooling and heating due to these species is also included.  The solution of the rate equations is carried out using a time-backwards differencing which insures stability; to maintain accuracy we sub-cycle the rate equations with a shorter timestep such that the electron and neutral fractions do not change by more than 10\%.

>>>>

\begin{equation}
\mathbb{F} = -\kappa_{\rm sp} \nabla T
\end{equation}

Where $\kappa_{\rm sp}$ is the Spitzer conduction coefficient
\citep{1962pfig.book.....S}.  Saturation of the heat flux in
high-temperature, low-density regimes (such as the intracluster medium
in galaxy clusters) is taken into account.  Thermal conduction in a
plasma can be strongly affected by the presence of magnetic field
lines, which may strongly suppress heat flow perpendicular to the
magnetic field.  In that case, we allow for heat transport only along
the magnetic field lines in simulations that include this physics, as follows:

\begin{equation}
\mathbb{F} = -\kappa_{\rm sp} \hat{b} (\hat{b} \cdot \nabla T)
\end{equation}

@@

In addition to this, \enzo\ includes the capability of following up to
12 chemical species using a non equilibrium-solver.  The species can
be turned on in sets, with the simplest model including just H, \Hp,
He, \Hep, \Hepp, e$^-$, and more complete models adding first species
important for gas-phase molecular hydrogen formation: \Hm, \HH~and
\HHp, and then HD formation: HD, D, \Dp.  The cooling and heating due
to these species is also included.  The solution of the rate equations
is carried out using a time-backwards differencing which insures
stability; to maintain accuracy we sub-cycle the rate equations with a
shorter timestep such that the electron and neutral fractions do not
change by more than 10\%.
